---
title: "PCA versus Factor Analysis"
subtitle: "Two methods that produce similar-looking output and answer different questions."
output:
  html_document:
    theme: null
    highlight: tango
    toc: true
    toc_depth: 2
    df_print: kable
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 9, fig.height = 6, dpi = 96
)
```

<!-- Web fonts load at view time; data-external keeps them out of the self-contained file -->
<link rel="stylesheet" data-external="1" href="https://fonts.googleapis.com/css2?family=Zilla+Slab:wght@500;600;700&family=Source+Sans+3:ital,wght@0,400;0,600;0,700;1,400&family=Source+Code+Pro:wght@400;600&display=swap">

## Purpose

Principal component analysis (PCA) and factor analysis (FA) are often confused, and in
practice they are sometimes used interchangeably. Both take a large set of correlated
variables and return a few dimensions with loadings and scores. But they rest on different
models of where the data come from. This walkthrough:

1. Explains the **conceptual and statistical difference** between FA and PCA.
2. Explains what `psych::principal()` is actually doing.
3. Fits a **four-dimension solution both ways** on the same census data and compares the
   loadings, the variance explained, and the scores.

It builds on the [Factor Analysis](measurement-factor-analysis.html) walkthrough, which uses
the same data and explains how the four-factor solution was chosen.

## 1. Two different questions

### Factor analysis

**Question:** *What latent constructs could be generating these observed variables?*

FA is a **measurement model**. Each observed variable is treated as an imperfect indicator of
a smaller set of latent factors, plus error:

$$
X = \Lambda F + \epsilon
$$

- $\Lambda$: loadings, how strongly each variable reflects each factor
- $\epsilon$: uniqueness, the variance not explained by the factors, including measurement
  error

FA models only the **shared (common) variance** and sets the unique variance aside.

### Principal component analysis

**Question:** *How can we re-express the data with fewer variables while keeping as much of
the variance as possible?*

PCA is a **data compression** method, a rotation of the coordinate system. Each component is
a weighted sum of the observed variables, chosen to capture as much variance as possible:

$$
C = W X
$$

The arrows run the other way. In FA, the factors *cause* the variables. In PCA, the
components are *built from* the variables. PCA does not model:

- latent causes
- measurement error
- the split between common and unique variance

> **One-line intuition.** FA tries to explain the *covariance* among variables. PCA tries to
> preserve their *total variance*.

## 2. Common versus total variance

| | Factor analysis | PCA |
|---|---|---|
| Models latent variables | Yes | No |
| Separates common from unique variance | Yes | No |
| Includes an error term | Yes | No |
| Treats all variance as signal | No | Yes |
| Built for theory and construct measurement | Yes | Sometimes |
| Built for compression and prediction features | Sometimes | Yes |

**The practical implication:** PCA loadings usually look *stronger* than FA loadings. PCA
counts all of a variable's variance, including the noise, as something to explain. FA
loadings reflect only the variance a variable shares with the others.

### What loadings and scores mean

**Loadings**

- **FA loading:** how strongly does this variable *measure* the latent factor?
- **PCA loading:** how strongly does this variable *contribute* to the component?

**Scores**

- **FA factor scores:** estimates of each case's value on the latent trait. They are
  model-based, and they contain estimation error.
- **PCA component scores:** exact projections of each case onto the component axes. They
  are geometry, not estimates.

Rotation, such as varimax, makes both solutions easier to read, but **rotating PCA does not
turn it into FA**.

## 3. What `psych::principal()` is doing

A common snippet looks like this:

```r
pc4 <- principal(d, nfactors = 4, rotate = "varimax", scores = TRUE)
print.psych(pc4, cut = 0.3, sort = TRUE)
```

This is **PCA**, even though the argument is named `nfactors`. `psych::principal()` extracts
principal components, optionally rotates them, and returns component scores. The `psych`
package uses the same argument names for both methods so the output is easy to compare, which
is also why the two are easy to confuse.

## 4. Setup and data

```{r packages}
# Install once if needed:
# install.packages(c("tidyverse", "psych", "GPArotation", "pander"))

library(tidyverse)     # data import and wrangling
library(psych)         # principal(), fa()
library(GPArotation)   # rotation methods
library(pander)        # clean tables in the knitted report

# Read local copies if they sit next to this file; otherwise read from the course repo.
repo <- "https://raw.githubusercontent.com/lecy/econometrics/main/lectures2/data/"
data_path <- if (file.exists("census-data.csv"))    "census-data.csv"    else paste0(repo, "census-data.csv")
dd_path   <- if (file.exists("census-data-dd.csv")) "census-data-dd.csv" else paste0(repo, "census-data-dd.csv")

out_dir <- "fa_vs_pca_outputs"
dir.create(out_dir, showWarnings = FALSE)
```

For comparability, both methods get identical inputs:

1. Median imputation for any missing values (a simple exploratory default)
2. Z-score standardization, so both methods analyze the correlation matrix

```{r prep}
dat <- readr::read_csv(data_path, show_col_types = FALSE)
dd  <- readr::read_csv(dd_path,   show_col_types = FALSE)

stopifnot(all(c("LABEL", "VARIABLE") %in% names(dd)))
vars <- dd$LABEL
stopifnot(all(vars %in% names(dat)))

median_impute <- function(x) ifelse(is.na(x), median(x, na.rm = TRUE), x)

X_z <- dat %>%
  dplyr::select(all_of(vars)) %>%
  mutate(across(everything(), ~ median_impute(as.numeric(.x)))) %>%
  scale() %>%
  as.data.frame()

dim(X_z)
```

## 5. Fit PCA and FA with four dimensions

Both use `varimax` rotation, so each solution has uncorrelated dimensions and is as easy to
read as possible.

### PCA with `psych::principal()`

```{r fit-pca}
pca4 <- psych::principal(
  X_z,
  nfactors = 4,           # number of components to keep
  rotate   = "varimax",   # orthogonal rotation
  scores   = TRUE         # return component scores
)
```

### FA with `psych::fa()`

Principal axis factoring (`fm = "pa"`) models only the common variance. As explained in the
Factor Analysis walkthrough, this four-factor solution produces a Heywood-case warning,
because two pairs of variables are near-duplicates. It is suppressed here.

```{r fit-fa4}
fa4 <- suppressWarnings(psych::fa(
  X_z,
  nfactors = 4,
  fm       = "pa",          # principal axis factoring
  rotate   = "varimax",
  scores   = "regression"   # regression-method factor scores
))
```

## 6. Line up the dimensions

Neither method guarantees that its first dimension matches the other method's first
dimension. PCA names its rotated components `RC1`–`RC4` and FA names its factors `PA1`–`PA4`,
and the numbering doesn't correspond. Before comparing, **match each factor to the component
whose scores correlate with it most strongly**, and flip the sign if needed. The sign of a
loading column is arbitrary in both methods.

```{r align}
cross <- cor(fa4$scores, pca4$scores)
round(cross, 2)

# For each FA factor, the PCA component it matches best
match_pc <- colnames(cross)[apply(abs(cross), 1, which.max)]
names(match_pc) <- rownames(cross)
stopifnot(!anyDuplicated(match_pc))
match_sign <- sign(cross[cbind(rownames(cross), match_pc)])
names(match_sign) <- rownames(cross)

# Construct names from the Factor Analysis walkthrough
factor_names <- c(PA1 = "ses", PA2 = "stability", PA4 = "immigrant", PA3 = "aging")
fa_order <- names(factor_names)

tibble(
  construct = factor_names[fa_order],
  fa_factor = fa_order,
  pca_component = match_pc[fa_order],
  score_correlation = round(cross[cbind(fa_order, match_pc[fa_order])], 3)
)
```

Every factor has a clear partner component. The four dimensions are the same constructs
under either method: socioeconomic status, household stability, immigrant communities, and
aging population.

## 7. Compare the loadings side by side

```{r compare-loadings}
L_fa  <- unclass(fa4$loadings)[, fa_order]
L_pca <- sweep(unclass(pca4$loadings)[, match_pc[fa_order]], 2, match_sign[fa_order], `*`)
colnames(L_fa)  <- paste0("FA_",  factor_names[fa_order])
colnames(L_pca) <- paste0("PCA_", factor_names[fa_order])

loadings_compare <- dd %>%
  dplyr::select(LABEL, VARIABLE) %>%
  left_join(as.data.frame(L_pca) %>% rownames_to_column("LABEL"), by = "LABEL") %>%
  left_join(as.data.frame(L_fa)  %>% rownames_to_column("LABEL"), by = "LABEL")

readr::write_csv(loadings_compare, file.path(out_dir, "loadings_pca_vs_fa.csv"))
```

Show each construct's PCA and FA loadings next to each other, for the variables where either
method has a loading of at least 0.40 on that construct.

```{r loadings-by-construct, results="asis"}
for (k in unname(factor_names[fa_order])) {
  tab <- loadings_compare %>%
    transmute(LABEL, VARIABLE,
              PCA = .data[[paste0("PCA_", k)]],
              FA  = .data[[paste0("FA_",  k)]],
              pca_minus_fa = abs(PCA) - abs(FA)) %>%   # compare sizes, ignoring sign
    filter(pmax(abs(PCA), abs(FA)) >= 0.40) %>%
    arrange(desc(abs(PCA))) %>%
    mutate(across(c(PCA, FA, pca_minus_fa), ~ sprintf("%.2f", .x)))
  cat("\n### ", k, "\n\n", sep = "")
  pander::pandoc.table(tab, justify = "llrrr", split.table = Inf)
}
```

The `pca_minus_fa` column compares the size of the two loadings, ignoring sign. Two patterns
stand out:

- **Strong indicators barely change.** College degrees, income, and homeownership load within
  about 0.03 of each other under both methods. When most of a variable's variance is shared,
  there is little unique variance for the two methods to disagree about.
- **Weaker indicators load higher in PCA.** Female labor force participation, children, and
  vacant units on the aging dimension, and percent Black on the stability dimension, gain
  roughly 0.08 to 0.12 under PCA. FA sets much of their variance aside as unique; PCA counts
  it. This is the common-versus-total-variance difference in action.

The main exceptions run the other way: percent speaking another language and percent 60 and
older load *higher* in FA. These are the near-duplicate variables behind the Heywood warning,
where FA treats essentially all of the variance as shared.

## 8. Variance explained and communalities

Both methods report a **communality** ($h^2$) for each variable: the share of its variance
captured by the four dimensions. In FA, the rest is uniqueness. In PCA, the rest is simply
variance the four components left behind.

```{r communalities}
comm <- tibble(
  LABEL   = names(fa4$communality),
  h2_pca  = unname(pca4$communality),
  h2_fa   = unname(fa4$communality)
) %>%
  left_join(dd, by = "LABEL") %>%
  relocate(LABEL, VARIABLE)

summarise(comm,
          mean_h2_pca = mean(h2_pca),
          mean_h2_fa  = mean(h2_fa),
          share_pca_higher = mean(h2_pca > h2_fa))

# Variables where FA reports the higher communality
comm %>% filter(h2_fa >= h2_pca) %>% mutate(across(c(h2_pca, h2_fa), ~ round(.x, 2)))
```

```{r communality-plot, fig.width=6.5, fig.height=6.5}
lim <- c(0, 1.05)
plot(comm$h2_fa, comm$h2_pca, xlim = lim, ylim = lim,
     pch = 19, col = "#bd5416", bty = "n",
     xlab = "Communality in factor analysis",
     ylab = "Communality in PCA",
     main = "Communality: PCA versus factor analysis")
abline(0, 1, lty = 2, col = "#6d7681")
text(comm$h2_fa, comm$h2_pca, comm$LABEL, pos = 4, cex = 0.6, col = "#4d565f")
```

Points above the dashed line are variables where PCA claims more variance than FA, which is
`r round(100 * mean(comm$h2_pca > comm$h2_fa))`% of them. The gap tends to be widest for
variables with low FA communalities, such as percent 17 and under and female labor force
participation. Those are the noisy indicators, where FA attributes most of the variance to
uniqueness and PCA keeps it. The few points below the line are the near-duplicates and a
handful of very strong indicators.

The same pattern shows up in total variance explained:

```{r variance-explained}
round(pca4$Vaccounted[c("SS loadings", "Proportion Var", "Cumulative Var"), ], 2)
round(fa4$Vaccounted[c("SS loadings", "Proportion Var", "Cumulative Var"), ], 2)
```

## 9. Compare the scores

Both methods produce a score on each dimension for every neighborhood. How similar are they,
and how independent is each set?

```{r compare-scores}
scores_fa <- as.data.frame(fa4$scores)[, fa_order]
names(scores_fa) <- paste0("FA_", factor_names[fa_order])

scores_pca <- sweep(as.data.frame(pca4$scores)[, match_pc[fa_order]], 2, match_sign[fa_order], `*`) %>%
  as.data.frame()
names(scores_pca) <- paste0("PCA_", factor_names[fa_order])

# Within-method correlations
round(cor(scores_pca), 3)
round(cor(scores_fa), 3)

# Matched pairs across methods
round(diag(cor(scores_pca, scores_fa)), 3)

readr::write_csv(scores_pca, file.path(out_dir, "scores_pca4.csv"))
readr::write_csv(scores_fa,  file.path(out_dir, "scores_fa4.csv"))
readr::write_csv(
  as.data.frame(round(cor(scores_pca, scores_fa), 3)) %>% rownames_to_column("PCA_component"),
  file.path(out_dir, "score_correlations_pca_vs_fa.csv")
)
```

Three things to notice:

- **PCA component scores are exactly uncorrelated.** They are exact projections onto
  orthogonal axes.
- **FA factor scores are only approximately uncorrelated.** The factors are orthogonal, but
  the scores are *estimates* of them, and estimation error introduces small correlations.
- **The matched scores are highly correlated** across methods (roughly 0.9 or higher). For a
  large, well-structured set of variables, the two methods agree on *which* dimensions exist.
  They disagree about how much of each variable is signal.

## 10. When to use which

### Prefer factor analysis when

- You want interpretable **latent constructs** that you can name and defend
- You care about **measurement validity**, meaning the split between common and unique
  variance
- You treat the variables as indicators of underlying neighborhood conditions

### Prefer PCA when

- You want compact features for **prediction** or **dimension reduction**
- You want to keep as much of the total variance as possible in fewer dimensions
- You don't need a latent-variable or measurement interpretation

### A practical workflow

Many projects:

1. Run PCA as a quick check of the structure.
2. Run FA to interpret and report the constructs.
3. Use FA factor scores, or PCA components, as predictors in later models.

## Outputs

Knitting this file writes four files to `fa_vs_pca_outputs/`:

- `loadings_pca_vs_fa.csv`: aligned PCA and FA loadings for every variable
- `scores_pca4.csv`: aligned PCA component scores
- `scores_fa4.csv`: FA factor scores
- `score_correlations_pca_vs_fa.csv`: correlations between the two sets of scores


```{css, echo=FALSE}
/* Matches the course notes pages (lectures2/*-notes.html) */
:root{
  --ground:#ffffff;
  --surface:#f3f4f6;
  --ink:#1f2429;
  --ink-mid:#4d565f;
  --ink-soft:#6d7681;
  --rule:#dcdfe4;
  --rule-strong:#c3c8cf;
  --accent:#bd5416;
  --accent-ink:#8f3f10;
  --accent-wash:#fbf0e7;
  --navy:#153243;
  --pumpkin:#dc965a;
  --font-display:"Zilla Slab",Georgia,"Times New Roman",serif;
  --font-body:"Source Sans 3","Segoe UI",Helvetica,Arial,sans-serif;
  --font-code:"Source Code Pro",Consolas,"Courier New",monospace;
  --measure:40rem;
}

*{box-sizing:border-box;}
html{-webkit-text-size-adjust:100%;}
body{
  margin:0;
  background:var(--ground);
  color:var(--ink);
  font-family:var(--font-body);
  font-size:1.0625rem;
  line-height:1.62;
  max-width:52rem;
  margin:0 auto;
  padding:0 1.5rem 5rem;
}

/* ---------- masthead (theme: null renders a bare h1.title + h3.subtitle) ---------- */
h1.title{
  font-family:var(--font-display);
  font-weight:700;
  font-size:2.75rem;
  line-height:1.06;
  letter-spacing:-.015em;
  margin:0;
  padding-top:3.5rem;
}
h1.title::before{
  content:"Measurement Theory · R walkthrough";
  display:block;
  font-family:var(--font-body);
  font-size:.8125rem;
  line-height:1.4;
  letter-spacing:.12em;
  text-transform:uppercase;
  font-weight:600;
  color:var(--accent-ink);
  margin-bottom:.9rem;
}
h3.subtitle{
  font-family:var(--font-body);
  font-weight:400;
  font-size:1.25rem;
  line-height:1.5;
  color:var(--ink-mid);
  margin:.9rem 0 0;
  padding-bottom:2rem;
  border-bottom:1px solid var(--rule);
  max-width:none;
}

/* ---------- table of contents ---------- */
#TOC{
  background:var(--surface);
  border-radius:10px;
  padding:1.25rem 1.5rem;
  margin:2rem 0 1rem;
}
#TOC::before{
  content:"Contents";
  display:block;
  font-size:.8125rem;
  letter-spacing:.14em;
  text-transform:uppercase;
  font-weight:700;
  color:var(--ink-soft);
  margin-bottom:.5rem;
}
#TOC ul{list-style:none;margin:0;padding:0;columns:2 16rem;column-gap:2rem;}
#TOC li{margin:.2rem 0;break-inside:avoid;}
#TOC ul ul{display:none;}
#TOC a{color:var(--ink);text-decoration:none;font-weight:600;font-size:.95rem;}
#TOC a:hover{color:var(--accent-ink);}

/* ---------- headings ---------- */
h2{
  font-family:var(--font-display);
  font-weight:600;
  font-size:2rem;
  line-height:1.14;
  margin:3.5rem 0 1rem;
  padding-top:1.5rem;
  border-top:1px solid var(--rule);
}
h3{
  font-family:var(--font-display);
  font-weight:600;
  font-size:1.45rem;
  line-height:1.24;
  margin:2.25rem 0 .75rem;
}
h4{font-family:var(--font-body);font-weight:700;font-size:1.05rem;margin:1.75rem 0 .5rem;}

p,ul,ol{max-width:var(--measure);}
p{margin:0 0 1.05rem;}
ul,ol{margin:0 0 1.05rem;padding-left:1.35rem;}
li{margin-bottom:.4rem;}
li::marker{color:var(--accent);font-weight:600;}
strong{font-weight:700;color:var(--ink);}
a{color:var(--accent-ink);text-underline-offset:2px;}
a:focus-visible{outline:2px solid var(--accent);outline-offset:2px;}
hr{border:0;border-top:1px solid var(--rule);margin:2.5rem 0;}

/* ---------- callouts (use > blockquotes) ---------- */
blockquote{
  background:var(--accent-wash);
  border:0;
  border-radius:3px;
  padding:1.1rem 1.3rem;
  margin:1.5rem 0;
  max-width:var(--measure);
  font-size:1rem;
}
blockquote p:last-child{margin-bottom:0;}

/* ---------- key-idea panel: wrap in <div class="concepts"> ---------- */
.concepts{
  background:var(--navy);
  color:#fff;
  border-radius:10px;
  padding:1.75rem 2rem 1.5rem;
  margin:1.5rem 0 2rem;
}
.concepts p,.concepts li{color:rgba(255,255,255,.9);}
.concepts strong{color:var(--pumpkin);}
.concepts li::marker{color:var(--pumpkin);}
.concepts code{background:rgba(255,255,255,.12);color:#fff;}

/* ---------- code ---------- */
code{font-family:var(--font-code);font-size:.88em;}
p code,li code,td code{background:var(--surface);padding:.08em .32em;border-radius:3px;color:var(--accent-ink);}
pre{
  font-family:var(--font-code);
  font-size:.85rem;
  line-height:1.5;
  background:var(--surface);
  border:0;
  border-radius:3px;
  padding:.9rem 1.1rem;
  margin:1rem 0 1.25rem;
  overflow-x:auto;
}
pre code{background:none;padding:0;color:inherit;font-size:inherit;}
pre.r{border-left:3px solid var(--accent);}
pre:not(.r){background:var(--ground);border:1px solid var(--rule);color:var(--ink-mid);}

/* ---------- math ---------- */
.math.display{display:block;overflow-x:auto;margin:1rem 0;}

/* ---------- tables ---------- */
table{
  border-collapse:collapse;
  font-size:.9rem;
  margin:1.25rem 0 1.75rem;
  display:block;
  max-width:100%;
  overflow-x:auto;
}
th,td{text-align:left;padding:.5rem .75rem;border-bottom:1px solid var(--rule);vertical-align:top;}
thead th{
  font-size:.75rem;
  letter-spacing:.09em;
  text-transform:uppercase;
  font-weight:700;
  color:var(--ink-soft);
  border-bottom:1.5px solid var(--rule-strong);
  white-space:nowrap;
}
td{font-variant-numeric:tabular-nums;}
caption{
  caption-side:top;
  text-align:left;
  font-size:.8rem;
  letter-spacing:.1em;
  text-transform:uppercase;
  font-weight:700;
  color:var(--ink-soft);
  padding-bottom:.6rem;
}

/* ---------- figures ---------- */
img{max-width:100%;height:auto;display:block;margin:1.25rem auto;}
p.caption{text-align:center;font-size:.9rem;color:var(--ink-soft);font-style:italic;}

@media (max-width:640px){
  h1.title{font-size:2.1rem;}
  h2{font-size:1.65rem;}
  h3{font-size:1.3rem;}
  body{padding:0 1rem 3.5rem;}
  .concepts{padding:1.35rem 1.25rem;}
}
```
