---
title: "Decomposition of Total Sum of Squares in Regression"
output:
  html_document:
    theme: readable
    highlight: tango
    toc: true
    self_contained: true
    number_sections: false
    css: textbook.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Sums of Squares in Regression

This document explains how the **Total Sum of Squares (TSS)** can be decomposed into the **Explained Sum of Squares (ESS)** and the **Residual Sum of Squares (RSS)** using the trick of inserting the regression line between the observed data \(y_i\) and the mean \(\bar{y}\).

# 1. Definitions

For observations \(y_i\), their mean \(\bar{y}\), and fitted values from a regression \(\hat{y}_i\):

- **Total Sum of Squares (TSS):**

$$
TSS = \sum_i (y_i - \bar{y})^2
$$

- **Explained Sum of Squares (ESS):**

$$
ESS = \sum_i (\hat{y}_i - \bar{y})^2
$$

- **Residual Sum of Squares (RSS):**

$$
RSS = \sum_i (y_i - \hat{y}_i)^2
$$

The decomposition is:

$$
TSS = ESS + RSS
$$

# 2. The Trick

We start with:

$$
y_i - \bar{y} = (y_i - \hat{y}_i) + (\hat{y}_i - \bar{y})
$$

Squaring both sides:

$$
(y_i - \bar{y})^2 = (y_i - \hat{y}_i)^2 + (\hat{y}_i - \bar{y})^2 + 2(y_i - \hat{y}_i)(\hat{y}_i - \bar{y})
$$

Summing over \(i\):

$$
TSS = RSS + ESS + 2\sum_i (y_i - \hat{y}_i)(\hat{y}_i - \bar{y})
$$

# 3. Orthogonality

In OLS regression, the cross-product vanishes:

$$
\sum_i (y_i - \hat{y}_i)(\hat{y}_i - \bar{y}) = 0
$$

Thus, we obtain the decomposition:

$$
TSS = ESS + RSS
$$

# 4. Example in R

Let’s demonstrate this with a small example using simulated data.

```{r example}
set.seed(123)

# Generate data
x <- 1:10
y <- 2 + 0.5 * x + rnorm(10, 0, 1)

# Fit regression
model <- lm(y ~ x)
yhat <- fitted(model)
ybar <- mean(y)

# Create components
df <- data.frame(
  X = x,
  Y = y,
  yhat = yhat,
  ybar = ybar,
  resid = y - yhat,
  fit_dev = yhat - ybar,
  total_dev = y - ybar
)

knitr::kable(df, digits=3)
```

# 5. Check Decomposition

We can see that:

$$
TSS = ESS + RSS
$$

```{r check}
TSS <- sum((df$total_dev)^2)
ESS <- sum((df$fit_dev)^2)
RSS <- sum((df$resid)^2)

c(TSS = TSS, ESS = ESS, RSS = RSS, ESS_plus_RSS = ESS + RSS) |> pander::pander()
```



